---
title: "Pair Programming - Data Science Lifecycle"
format: html
editor: visual
bibliography: references.bib
execute:
  echo: true
  message: false
  warning: false
editor_options: 
  chunk_output_type: console
---

# Pair Programming

In pair programming exercises learners work in pairs to solve exercises together.

# Exercise 2 : Data visualisation

Work through the next two tasks by yourself.

## Task 1: Visualise penguins data

1.  Remove the hashtags to uncomment the lines in the code-chunk below with the label "penguins-vis".
2.  Run the code that is contained in the code-chunk.
3.  Render the file.

```{r}

# ggplot(data = penguins, 
#         mapping = aes(x = bill_depth_mm, 
#                       y = bill_length_mm, 
#                       color = species)) +
#    geom_point()

```

## Task 2: Visualise waste data

1.  Fill in the blanks in the code-chunk below.
2.  Run the code that is contained in the code-chunk.
3.  Render the file. Why does the plot not show up?

```{r}
#| eval: false

ggplot(data = ___,
       mapping = ___(x = Category, 
                     y = Mass)) +
  ____col() +
  coord_flip() 

```

## Task 3: Visualise gapminder data

1.  Run the code that is contained in the code-chunk below.
2.  Render the file and view the output.
3.  Describe what the function `facet_wrap()` does.

```{r}

ggplot(data = gapminder,
       mapping = aes(x = year,
                     y = lifeExp,
                     group = country)) +
  geom_line() +
  facet_wrap(~continent)

```

# References
